Linear vertical shear instability in astrophysical disks: a numerical study

### 2014-11-28 linear_iso.tex
minor update

\subsection{Stability in the absence of vertical shear}
When $q=0$, the governing equation is
\begin{align}
  \frac{d}{d\hat{z}}\left[w(\hat{z})\frac{dV}{d\hat{z}}\right] + nV
  w(\hat{z}) =0, 
\end{align}
i.e. Hermite's differential equation. For bounded solutions we have 
\begin{align}
  V\propto \He_n(\hat{z}),
\end{align}
and $n$ is a non-negative integer. This translates to
\begin{align}
  \left|\frac{\sigma}{\Omega_k}\right| = \sqrt{n}
  \left(1+\hat{k}^2\right)^{-1/2}. 
\end{align}
Since we have assumed $|\sigma^2|\ll \kappa^2$, our analysis is only
valid for large wavenumbers such that $\hat{k}^2\gg 
n-1$. Physically, this corresponds to radial length-scales much
smaller than the local disk scale height. From

\subsection{Instability in the presence of weak vertical shear}
To investigate the destabilizing effect of vertical shear, we first
consider a system with $q\equiv0$, for which the eigenfunctions are
Hermite polynomials and the real eigenfrequencies are known. We then
perturb this system by introducing a weak vertical shear ($|q|\ll1$)
and linearize the governing equation. This procedure is
\begin{align}   
  &q \to 0 + \delta q\\
  &V \to \He_n + \delta V \\
  &\sigma \to \sigma + \delta\sigma, 
\end{align}
where for this exercise we restore $\sigma$ as the eigenfrequency.

Linearizing the integral Eq. \ref{integral_relation} and taking the
imaginary part, we have
\begin{align}
  \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
  \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
  \He_n^2(\hat{z}) d\hat{z} \notag\\
  = \delta q h \hat{k} 
  \int_{-\infty}^{\infty}
  w(\zhat)\hat{z}\He_n(\hat{z})\frac{d\He_n}{d\zhat} d\hat{z}
\end{align}
Integrating the right hand side by parts and recognizing $\He_2 =
\hat{z}^2 -1$, we have
\begin{align}
  \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
  \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
  \He_n^2(\hat{z}) d\hat{z} \notag\\
  = \frac{1}{2}\delta q h \hat{k} 
  \int_{-\infty}^{\infty}
  w(\zhat)\He_2(\hat{z})\He_n^2(\hat{z})d\hat{z}. 
\end{align}
Finally, we evaluate the integrals using following properties of Hermite polynomials
\begin{align}
  \int_{-\infty}^{\infty}
  w(\xi) \He_k(\xi)\He_l(\xi) d\xi = \sqrt{2\pi}k!\delta_{kl}, 
\end{align}
and
\begin{align}
  \int_{-\infty}^{\infty}
  w(\xi) \He_m(\xi)\He_k(\xi)\He_l(\xi) d\xi =
  \frac{\sqrt{2\pi}m!k!l!}{(s-m)!(s-k)!(s-l)!}, 
\end{align}
where $s = (m+k+l)/2$. We then obtain
\begin{align}
 \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
 \left(1+\hat{k}^2\right) = \delta q h \hat{k} \frac{n!}{(n-1)!}.
\end{align}
Inserting the original eigenfrequency $\sigma = \pm
\sqrt{n}\Omega_k(1+\hat{k}^2)^{-1/2}$ gives
\begin{align}
  \frac{\imag(\delta\sigma)}{\Omega_k} = \pm \frac{1}{2}\delta q h
  \sqrt{n} \frac{\hat{k}}{\sqrt{1+\hat{k}^2}}. 
\end{align}
 For $\hat{k}\gg 1$ we have
\begin{align}
  \frac{\imag(\delta\sigma)}{\Omega_k} \simeq \pm \frac{1}{2}\delta q h
  \sqrt{n} \sgn{\hat{k}}. 
\end{align}

>>>>

\subsection{Stability in the absence of vertical shear}
When $q=0$, the governing equation is
\begin{align}
  \frac{d}{d\hat{z}}\left[w(\hat{z})\frac{dV}{d\hat{z}}\right] + nV
  w(\hat{z}) =0, 
\end{align}
i.e. Hermite's differential equation. If we impose that the kinetic
energy density remain bounded at infinity, then  
\begin{align}
  V\propto \He_n(\hat{z}),
\end{align}
and $n$ is a non-negative integer. This translates to a real
eigenfrequency $\sigma$ such that
\begin{align}
  \left|\frac{\sigma}{\Omega_k}\right| = \sqrt{n}
  \left(1+\hat{k}^2\right)^{-1/2}. 
\end{align}
Since we have assumed $|\sigma^2|\ll \kappa^2$, our analysis is only
valid for large wavenumbers such that $\hat{k}^2\gg 
n-1$. Physically, this corresponds to radial length-scales much
smaller than the local disk scale height. From

\subsection{Simple solutions with vertical shear}
The differential equation, Eq. \ref{iso_ode3}, can be solved by
\begin{align}
  Q\propto \hat{z}. 
\end{align} 
%Although $Q$ diverges at infinity, this solution is
%acceptable because the implied kinetic energy densities remain
%bounded, since the density decays rapidly. 
In this case the eigenfrequency must satisfy 
\begin{align}
  \left(\frac{\sigma}{\Omega_k}\right)^2 = \frac{1+\ii q h
    \hat{k}}{1+\hat{k}^2}, 
\end{align}
and $\sigma$ is generally complex\footnote{
We note that an even simpler solution to Eq. \ref{iso_ode3} is
$Q=\mathrm{constant}$, but this implies $\sigma^2=0$ and therefore
stability, so this case is of no interest.}, with  
%If we write $\sigma =
%(\hat{\omega} + \ii \hat{\nu})\Omega_k$ with $\hat{\omega},\hat{\nu}$
%real, we have
\begin{align}
  \left[\frac{\imag(\delta\sigma)}{\Omega_k}\right]^2 =
  \frac{\sqrt{1+q^2h^2\hat{k}^2} - 1}{2\left(1+\hat{k}^2\right)}. 
\end{align}
This expression can be simplified in the limit $|q|\ll 1$. Then we have
\begin{align}\label{simple_growth}
  \frac{\imag(\sigma)}{\Omega_k} = \pm
  \frac{qh\hat{k}}{2\sqrt{1+\hat{k}^2}} \quad\quad (|q|\ll 1).  
\end{align}

\subsection{Instability in the presence of weak vertical shear}
We can also investigate the destabilizing effect of vertical shear
without explicitly solving the full governing equation,
Eq. \ref{iso_ode3}. We first
consider a system with $q\equiv0$, for which the eigenfunctions are
Hermite polynomials and the real eigenfrequencies are known. We then
perturb this system by introducing a weak vertical shear ($|q|\ll1$)
and linearize the governing equation. This procedure is
\begin{align}   
  &q \to 0 + \delta q\\
  &V \to \He_n + \delta V \\
  &\sigma \to \sigma + \delta\sigma, 
\end{align}
where for this exercise we restore $\sigma$ as the eigenfrequency.

Linearizing the integral Eq. \ref{integral_relation} and taking the
imaginary part, we have
\begin{align}
  \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
  \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
  \He_n^2(\hat{z}) d\hat{z} \notag\\
  = \delta q h \hat{k} 
  \int_{-\infty}^{\infty}
  w(\zhat)\hat{z}\He_n(\hat{z})\frac{d\He_n}{d\zhat} d\hat{z}
\end{align}
Recognizing $\hat{z} = \He_1$ and using $d\He_n/d\hat{z} = n
\He_{n-1}$ we have
 \begin{align}
   \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
   \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
   \He_n^2(\hat{z}) d\hat{z} \notag\\
   =\delta q h \hat{k} 
   \int_{-\infty}^{\infty}
   w(\zhat)\He_1(\hat{z})\He_n(\hat{z})n\He_{n-1}(\hat{z})d\hat{z}. 
 \end{align}
% Integrating the right hand side by parts and recognizing $\He_2 =
% \hat{z}^2 -1$, we have
% \begin{align}
%   \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
%   \left(1+\hat{k}^2\right) \int_{-\infty}^{\infty} w(\zhat)
%   \He_n^2(\hat{z}) d\hat{z} \notag\\
%   = \frac{1}{2}\delta q h \hat{k} 
%   \int_{-\infty}^{\infty}
%   w(\zhat)\He_2(\hat{z})\He_n^2(\hat{z})d\hat{z}. 
% \end{align}
Finally, we evaluate the integrals using following properties of Hermite polynomials
\begin{align}
  \int_{-\infty}^{\infty}
  w(\xi) \He_k(\xi)\He_l(\xi) d\xi = \sqrt{2\pi}k!\delta_{kl}, 
\end{align}
and
\begin{align}
  \int_{-\infty}^{\infty}
  w(\xi) \He_m(\xi)\He_k(\xi)\He_l(\xi) d\xi =
  \frac{\sqrt{2\pi}m!k!l!}{(s-m)!(s-k)!(s-l)!}, 
\end{align}
where $s = (m+k+l)/2$. We then obtain
% \begin{align}
%  \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
%  \left(1+\hat{k}^2\right) = \delta q h \hat{k} \frac{n!}{(n-1)!}.
% \end{align
 \begin{align}
  \frac{2\sigma\imag(\delta\sigma)}{\Omega_k^2}
  \left(1+\hat{k}^2\right) = \delta q h \hat{k} n.
 \end{align}
Inserting the original eigenfrequency $\sigma = \pm
\sqrt{n}\Omega_k(1+\hat{k}^2)^{-1/2}$ gives
\begin{align}
  \frac{\imag(\delta\sigma)}{\Omega_k} = \pm \frac{1}{2}\delta q h
  \sqrt{n} \frac{\hat{k}}{\sqrt{1+\hat{k}^2}}. 
\end{align}
This result agrees with Eq. \ref{simple_growth} for $n=1$, since the
function $\He_1$ solves the governing equation exactly. For $\hat{k}\gg 1$ we have
\begin{align}
  \frac{\imag(\delta\sigma)}{\Omega_k} \simeq \pm \frac{1}{2}\delta q h
  \sqrt{n} \sgn{\hat{k}}. 
\end{align}
